Box

### 2021-07-16 MS/box.Rmd
Minor fixes from Megan

Here we show how to obtain sampling variance for a slope (or a 'rate of change') when you are given measurements of a physiological trait at two points in an environmental gradient (e.g. temperature, salinity, acidity). We also go on to derive sampling variance for the difference between two slopes, and finally end on showing how SMD can be corrected to account for both unitary and temperature differences across effect sizes. The sampling variances associated with these 'new' effect sizes will allow one to take advantage of powerful meta-analytic models.

@@

It may be the best to start with a real example that comparative physiologists can easily associate, namely the temperature acclimation slope or as we call it the (temperature) acclimation response ratio (ARR). ARR can be defined as a slope for acclimated physiological responses at two different temperature points, defined as:

@@

where *T* stands for temperature in degrees Celsius and $T_{2} > T_{1}$, and $M_{1}$ and $M_{2}$ are the average physiological responses (estimated means) at two temperature points $T_{1}$ and $T_{2}$.

@@

Importantly, when $M_{1}$ and $M_{2}$ are independent of each other the covariance is 0. More concretely, if measurements are done on two different groups of animals at two different temperature points ($T_1$ and $T_2$), then the co-variances between these two sets of measurements are 0.

@@

Now, it is not uncommon that we use the same group of animals (or plants) at two temperature points. If we do, however, then we need to add the co-variance between $M_{1}$ and $M_{2}$ ($\text{Cov}(M_{1},M_{2})$); note that as above, the co-variance equals the corresponding correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance can be now written as:

@@

By assuming the number of animals (*N*) are the same at the two temperature points, we can slightly simplify this formula:

@@

where *r* is the correlation between a set of measurement $M_1$ and $M_2$ form the same individuals at two point $T_1$ and $T_2$. As you may notice, we need raw data to calculate *r*. So, in reality, we often need to assume a certain value of *r*. @Noble2017 discuss why we can assume $r = 0.5$ and this may be the most reasonable thing to do when we do not have any estimate of *r*.

@@

Now let us assume that we now want to know the difference between two different ARR values: female ARR, $\text{ARR}_{f}$ and Male ARR, $\text{ARR}_{m}$. Such a difference (ARRD) can be written as:

@@

Now we have derive sampling variances for 'temperature' ARR and ARRD for both independent and dependent cases. But these formulas can be applied to changes in salinity, PH, oxygen and other abotic factors. Also, the constant components of these formulas can be flexibly adjusted. For example, it may be possible that males and females were measured in slightly different temperatures; say females at $T_3$ and $T_4$ ($T_4 > T_3$) and male at $T_5$ and $T_6$ ($T_6 > T_5$) like below:

@@

Also, note that the difference between slopes does not need to be that of males and females. It can be between experimental and control groups or comparing any two groups or situations.

@@

ARR assumes that the units for the mean difference across studies is consistent. This will often not be the case for many meta-analyses that hope to synthesise a wide diversity of trait types that vary in their units (e.g., ul/g, grams, seconds, minutes). Fortunately, $lnRR$ and SMD (Table 1) already control for unitary differences. As such, we can apply some of the same logic we discuss above to SMD to correct for both temperature and unitary differences. Table 1 provides the formula for SMD, it's pooled standard deviation, and sampling variance. We won't re-iterate these formulas again. Just like with ARR we can apply our temperature correction to the SMD formula as follows:

@@

Here, we can see that the difference between $M_1$ and $M_2$ is standardised by the pooled SD, essentially correcting for unitary differences. Dividing SMD by the temperature difference on top of that results in correcting the effect size further by temperature. To derive the sampling variance for $SMD_T$ we can apply the same principles as we did above to derive an approximate sampling variance as:

@@

where $f(M_1)$ represents the function of the random variable $X_1$, and importantly, $f'(M_1)$ is the first derivative of $f(M_1)$. Probably it is hard to understand what does this really means without a concrete example. So let's derive the sampling variance for $\ln{RR}$, which is:

@@

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}}\right ) * \left(\frac{1}{M_{1}} \right)^2 + \left ( \frac{SD_{2}^2}{N_{2}}\right ) * \left(\frac{1}{M_{2}} \right)^2 =   \left ( \frac{SD_{1}^2}{N_{1}M_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}M_{2}^2}\right )
$$
Note here again that the first term, $\frac{SD_{1}^2}{N_{1}}$ is the sampling variance for group 1, and we simply multiply this term by the square of the first derivative of $M_{1}$ as the Delta method tells us. Of course, readers will see that this is equivalent to the sampling variance for $ln_{RR}$ (Table 1). We can extend this sampling variance to include dependency between the groups as:

>>>>

Here we show how to obtain sampling variance for a slope (or a 'rate of change') when you have measurements of a physiological trait at two points along an environmental gradient (e.g. temperature, salinity, acidity). We also go on to derive sampling variance for the difference between two slopes, and finally end on showing how SMD can be corrected to account for both unitary and temperature differences across effect sizes. The sampling variances associated with these 'new' effect sizes will allow one to take advantage of powerful meta-analytic models.

@@

Let's start with a real example that comparative physiologists can easily associate with, namely the temperature acclimation slope or as we call it the (temperature) acclimation response ratio (ARR). ARR can be defined as a slope for acclimated physiological responses at two different temperature points, defined as:

@@

where *T* stands for temperature in degrees Celsius and $T_{2} > T_{1}$, and $M_{1}$ and $M_{2}$ are the average physiological responses (estimated means) at temperature points $T_{1}$ and $T_{2}$.

@@

Importantly, when $M_{1}$ and $M_{2}$ are independent of each other the covariance is 0. More concretely, if measurements are taken on two different groups of animals at two different temperatures ($T_1$ and $T_2$), then the co-variances between these two sets of measurements are 0.

@@

Now, it is not uncommon that we measure the same group of animals (or plants) at two temperatures. If we do, however, then we need to add the co-variance between $M_{1}$ and $M_{2}$ ($\text{Cov}(M_{1},M_{2})$); note that as above, the co-variance equals the corresponding correlation multiplied by the square-root of two (sampling) variances. Therefore, the sampling variance can be now written as:

@@

By assuming the number of animals (*N*) are the same at the two temperatures, we can slightly simplify this formula:

@@

where *r* is the correlation between a set of measurement $M_1$ and $M_2$ form the same individuals at two point $T_1$ and $T_2$. As you may notice, we need raw data to calculate *r*. So, in reality, we often need to assume a certain value of *r*. When we do not have an estimate of *r* we can reasonably assume it to be 0.5 [@Noble2017].

@@

Now let us assume that we want to know the difference between two different ARR values: female ARR, $\text{ARR}_{f}$ and Male ARR, $\text{ARR}_{m}$. Such a difference (ARRD) can be written as:

@@

Here we have derived sampling variances for 'temperature' ARR and ARRD for both independent and dependent cases. But these formulas can also be applied to changes in other factors (e.g., salinity, pH, oxygen). Furthermore, the constant components of these formulas can be flexibly adjusted. For example, it may be possible that males and females were measured in slightly different temperatures; say females at $T_3$ and $T_4$ ($T_4 > T_3$) and male at $T_5$ and $T_6$ ($T_6 > T_5$) like below:

@@

Also, note that the difference between slopes does not need to be that of males and females. This formula can be used to compare any two groups or situations.

@@

ARR assumes that the units for the mean difference across studies is consistent. This will often not be the case for many meta-analyses that hope to synthesise a wide array of trait types that vary in their units (e.g., ul/g, grams, seconds, minutes). Fortunately, $lnRR$ and SMD (Table 1) already control for unitary differences. As such, we can apply some of the same logic we discuss above to SMD to correct for both temperature and unitary differences. Table 1 provides the formula for SMD, it's pooled standard deviation, and sampling variance. We won't re-iterate these formulas again. Just like with ARR we can apply our temperature correction to the SMD formula as follows:

@@

Here, we can see that the difference between $M_1$ and $M_2$ is standardised by the pooled SD, essentially correcting for unitary differences. Additionally, dividing SMD by the temperature difference results in further correcting the effect size by temperature. To derive the sampling variance for $SMD_T$ we can apply the same principles as we did above to derive an approximate sampling variance as:

@@

where $f(M_1)$ represents the function of the random variable $X_1$, and importantly, $f'(M_1)$ is the first derivative of $f(M_1)$. Let's demonstrate with a concrete example by re-deriving the sampling variance for $\ln{RR}$, which is:

@@

$$
\sigma^2_{ln{RR}} \approx  \left ( \frac{SD_{1}^2}{N_{1}}\right ) * \left(\frac{1}{M_{1}} \right)^2 + \left ( \frac{SD_{2}^2}{N_{2}}\right ) * \left(\frac{1}{M_{2}} \right)^2 =   \left ( \frac{SD_{1}^2}{N_{1}M_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}M_{2}^2}\right )
$$
Note here again that the first term, $\frac{SD_{1}^2}{N_{1}}$ is the sampling variance for group 1, and we simply multiply this term by the square of the first derivative of $M_{1}$ as the Delta method tells us. Readers may notice that this is equivalent to the sampling variance for $ln_{RR}$ (Table 1). We can extend this sampling variance to include dependency between the groups as:
